---
title: "Strava Heatmap"
date: "2025-11-02"
description: "Creating a Strava Heatmap with dowloaded activity files using ggpmap in R"
categories: [mapping, R, Strava, heatmap, ggmap]
author: Marc Weber
---

I wanted to create a basic heatmap like you can create based on your activities in Strava. The challenge was downloading / scraping all the data, wrangling it into a format to use, adding an underlying basemap from Stamen and mapping it using ggpmap in R. I bulk downloaded the activies from my Strava account page and then processed in R. I also generated an interactive heatmap with leaflet.

### Load Libraries

```{r}
#| warning: false
library(tidyverse)
library(sf)
library(FITfileR)
library(R.utils)
library(ggplot2)
library(ggmap)
library(leaflet)
library(leaflet.extras)
library(arrow)
```

### Uncompress the downloaded activity data

```{r}
#| warning: false
activities_path <- "C:/Users/mwebe/OneDrive/GitProjects/mhweber.github.io/posts/2025-11-01-Strava-R-map/strava_download/activities"

# gpx_files <- list.files(path = activities_path, pattern = "*.fit.gz", full.names = TRUE)
# 
# 
# # Loop through each .gz file and decompress it
# for (file_path in gpx_files) {
#   base_filename <- basename(file_path)
#   unzipped_filename <- sub("\\.gz$", "", base_filename)
#   gunzip(file_path, remove = FALSE)
# }

fit_files <- list.files(path = activities_path, full.names = TRUE)
```

### Create a function to extract data from all the .fit files

```{r}
#| warning: false
process_fit_files <- function(file_path) {
  if (!file.exists(file_path)) {
    warning(paste("File not found:", file_path))
    return(NULL)
  }

  tryCatch({
    # Read the FIT file
    fit_file <- readFitFile(file_path)

    # Extract 'record' messages, which contain GPS data
    records_list <- records(fit_file)

    if (is.null(records_list) || (is.list(records_list) && length(records_list) == 0)) {
      message(paste("No 'record' messages found in", basename(file_path)))
      return(NULL)
    }

    # Combine potential multiple tibbles (due to different message definitions) into one
    # If it's a single tibble, it's put in a list for consistent processing
    if (is_tibble(records_list)) {
        combined_records <- records_list
    } else {
        combined_records <- bind_rows(records_list)
    }

    # Select key coordinate information and convert to a tibble
    coordinates_data <- combined_records %>%
      select(
        timestamp,
        position_lat,
        position_long,
        # Optional fields, include if they exist
        if ("enhanced_altitude" %in% names(combined_records)) "enhanced_altitude" else if ("altitude" %in% names(combined_records)) "altitude" else NULL,
        if ("distance" %in% names(combined_records)) "distance" else NULL
      ) %>%
      # Remove rows where location data is missing
      na.omit()

    return(coordinates_data)

  }, error = function(e) {
    warning(paste("Error processing file", file_path, ":", e$message))
    return(NULL)
  })
}
```

### Run the function to create a dataframe of all activities

This can take a while if you have many activities

```{r}
#| eval: false
#| warning: false
all_data <- map_dfr(fit_files, process_fit_files)
```

```{r}
#| eval: false
#| warning: false
#| echo: false
write_parquet(all_data,'C:/Users/mwebe/OneDrive/GitProjects/mhweber.github.io/posts/2025-11-01-Strava-R-map/all_data.parquet')
```

```{r}
#| warning: false
#| echo: false
# View the combined data
all_data <- read_parquet('C:/Users/mwebe/OneDrive/GitProjects/mhweber.github.io/posts/2025-11-01-Strava-R-map/all_data.parquet')
all_data$timestamp <- as.POSIXct(all_data$timestamp, format = "%Y-%m-%d %H:%M:%S")

all_data <- all_data  |> 
      filter(timestamp > as.POSIXct("2023-01-01 00:00:00", format = "%Y-%m-%d %H:%M:%S"))

head(all_data)
```

### Truncate the data to just local activities

```{r}
#| warning: false
filtered_df <- all_data |> 
  filter(position_lat > 44.4 & position_lat < 44.8) |> 
  filter(position_long > -123.6 & position_long < -123.1)
```

### Bin the data for better viewing in ggmap

```{r}
#| warning: false
binned_data <- filtered_df %>%
  mutate(
    lon_bin = cut(position_long, breaks = 75),
    lat_bin = cut(position_lat, breaks = 75)
  ) %>%
  group_by(lon_bin, lat_bin) %>%
  summarise(count = n(), .groups = 'drop') %>%
  # Convert bin factors back to numeric midpoints for plotting
  mutate(
    lon_mid = as.numeric(sub(',.*', '', sub('\\(', '', lon_bin))) + (as.numeric(sub('.*,', '', sub('\\]', '', lon_bin))) - as.numeric(sub(',.*', '', sub('\\(', '', lon_bin))))/2,
    lat_mid = as.numeric(sub(',.*', '', sub('\\(', '', lat_bin))) + (as.numeric(sub('.*,', '', sub('\\]', '', lat_bin))) - as.numeric(sub(',.*', '', sub('\\(', '', lat_bin))))/2
  )
```

### Get Stamen background map

```{r}
#| warning: false
# Get the map tiles
map_bbox <-c(left = -123.5, bottom = 44.4, right = -123, top = 44.9)
base_map <- get_stadiamap(bbox = map_bbox, zoom = 12, maptype = "stamen_terrain_lines")
```

## Make the static map with ggmap

```{r}
#| warning: false
ggmap(base_map) +
  geom_tile(data=binned_data, aes(x = lon_mid, y = lat_mid, fill = count)) +
  scale_fill_gradientn(colors = c("blue", "green", "yellow", "red")) + # Customize color scale
  theme_minimal() +
  labs(title = "Strava Activity Density Heatmap", fill = "Activity Count") +
  coord_fixed(ratio = 1.3)
```

### Make an interactive heatmap with leaflet

```{r}
#| warning: false
run_map <- filtered_df  |> 
  leaflet()  |> 
  addTiles()  |>  # Add default OpenStreetMap map tiles
  addHeatmap(
    lng = ~position_long,
    lat = ~position_lat,
    blur = 20,    # Adjust blur and radius for desired look
    max = 0.5,
    radius = 15,
  )

# Display the map
run_map

```
