---
title: "Strava Heatmap"
date: "2025-11-02"
description: "Creating a Strava Heatmap with dowloaded activity files using ggpmap in R"
categories: [mapping, R, Strava, heatmap, ggmap]
author: Marc Weber
---

I wanted to create a basic heatmap like you can create based on your activities in Strava. The challenge was downloading / scraping all the data, wrangling it into a format to use, adding an underlying basemap from Stamen and mapping it using ggpmap in R. I bulk downloaded the activies from my Strava account page and then processed in R. I also generated an interactive heatmap with leaflet.

### Load Libraries

```{r}
#| warning: false
library(tidyverse)
library(sf)
library(tigris)
library(FITfileR)
library(litedown)
library(R.utils)
library(ggplot2)
library(osmdata)
library(ggmap)
library(leaflet)
library(leaflet.extras)
library(arrow)
```

### Uncompress the downloaded activity data

```{r}
#| warning: false
activities_path <- "C:/Users/mwebe/OneDrive/GitProjects/mhweber.github.io/posts/2025-11-01-Strava-R-map/strava_download/activities"

# gpx_files <- list.files(path = activities_path, pattern = "*.fit.gz", full.names = TRUE)
# 
# 
# # Loop through each .gz file and decompress it
# for (file_path in gpx_files) {
#   base_filename <- basename(file_path)
#   unzipped_filename <- sub("\\.gz$", "", base_filename)
#   gunzip(file_path, remove = FALSE)
# }

fit_files <- list.files(path = activities_path, full.names = TRUE)
```

### Create a function to extract data from all the .fit files

```{r}
#| warning: false
process_fit_files <- function(file_path) {
  if (!file.exists(file_path)) {
    warning(paste("File not found:", file_path))
    return(NULL)
  }

  tryCatch({
    # Read the FIT file
    fit_file <- readFitFile(file_path)

    # Extract 'record' messages, which contain GPS data
    records_list <- records(fit_file)

    if (is.null(records_list) || (is.list(records_list) && length(records_list) == 0)) {
      message(paste("No 'record' messages found in", basename(file_path)))
      return(NULL)
    }

    # Combine potential multiple tibbles (due to different message definitions) into one
    # If it's a single tibble, it's put in a list for consistent processing
    if (is_tibble(records_list)) {
        combined_records <- records_list
    } else {
        combined_records <- bind_rows(records_list)
    }

    # Select key coordinate information and convert to a tibble
    coordinates_data <- combined_records %>%
      select(
        timestamp,
        position_lat,
        position_long,
        # Optional fields, include if they exist
        if ("enhanced_altitude" %in% names(combined_records)) "enhanced_altitude" else if ("altitude" %in% names(combined_records)) "altitude" else NULL,
        if ("distance" %in% names(combined_records)) "distance" else NULL
      ) %>%
      # Remove rows where location data is missing
      na.omit()

    return(coordinates_data)

  }, error = function(e) {
    warning(paste("Error processing file", file_path, ":", e$message))
    return(NULL)
  })
}
```

### Run the function to create a dataframe of all activities

This can take a while if you have many activities

```{r}
#| eval: false
#| warning: false
all_data <- map_dfr(fit_files, process_fit_files)
```

```{r}
#| eval: false
#| warning: false
#| echo: false
write_parquet(all_data,'C:/Users/mwebe/OneDrive/GitProjects/mhweber.github.io/posts/2025-11-01-Strava-R-map/all_data.parquet')
```

```{r}
#| warning: false
#| echo: false
# View the combined data
all_data <- read_parquet('C:/Users/mwebe/OneDrive/GitProjects/mhweber.github.io/posts/2025-11-01-Strava-R-map/all_data.parquet')
all_data$timestamp <- as.POSIXct(all_data$timestamp, format = "%Y-%m-%d %H:%M:%S")

head(all_data)
```

### Truncate the data to just local activities within the last year and make spatial

```{r}
#| warning: false
filtered_df <- all_data |> 
  filter(position_lat > 44.5 & position_lat < 44.7) |> 
  filter(position_long > -123.4 & position_long < -123.2) |> 
  filter(timestamp > as.POSIXct("2025-01-01 00:00:00", format = "%Y-%m-%d %H:%M:%S"))
filtered_df_sf <- st_as_sf(filtered_df, coords = c('position_long', 'position_lat'))  |>  st_set_crs(4326)

```

### Bin the data for better viewing in ggmap

```{r}
#| warning: false
binned_data <- filtered_df %>%
  mutate(
    lon_bin = cut(position_long, breaks = 75),
    lat_bin = cut(position_lat, breaks = 75)
  ) %>%
  group_by(lon_bin, lat_bin) %>%
  summarise(count = n(), .groups = 'drop') %>%
  # Convert bin factors back to numeric midpoints for plotting
  mutate(
    lon_mid = as.numeric(sub(',.*', '', sub('\\(', '', lon_bin))) + (as.numeric(sub('.*,', '', sub('\\]', '', lon_bin))) - as.numeric(sub(',.*', '', sub('\\(', '', lon_bin))))/2,
    lat_mid = as.numeric(sub(',.*', '', sub('\\(', '', lat_bin))) + (as.numeric(sub('.*,', '', sub('\\]', '', lat_bin))) - as.numeric(sub(',.*', '', sub('\\(', '', lat_bin))))/2
  )

binned_data_sf <- st_as_sf(binned_data, coords = c('lon_mid', 'lat_mid'))  |>  st_set_crs(4326)
```

### Get Stamen background map

```{r}
#| warning: false
# Get the map tiles
map_bbox <-c(left = -123.4, bottom = 44.5, right = -123.2, top = 44.7)
base_map <- get_stadiamap(bbox = map_bbox, zoom = 12, maptype = "stamen_terrain_lines")
```

### Get background roads as an sf object
First query for hiking trails within the bounding box of activity data.  Then convert the osm data to an `sf` object and extract just the trail line features
```{r}
roads <- roads("OR", c("Benton","Polk"), progress_bar = FALSE)
roads <- roads |> 
  st_transform(roads,crs=st_crs(filtered_df_sf)) |> 
  st_crop(st_bbox(filtered_df_sf))
```

## Make a static map with ggmap

```{r}
#| warning: false
ggmap(base_map) +
  geom_tile(data=binned_data, aes(x = lon_mid, y = lat_mid, fill = count)) +
  scale_fill_gradientn(colors = c("blue", "green", "yellow", "red")) + # Customize color scale
  theme_minimal() +
  labs(title = "Strava Activity Density Heatmap", fill = "Activity Count") +
  coord_fixed(ratio = 1.3)
```

### A more pleasing static map with ggplot and cowplot
Inspired by great example of mapping Strava data in [this great R-bloggers post](https://www.r-bloggers.com/2021/03/heatmapping-my-new-york-city-marathon-training/)

```{r}
#| warning: false
p <- ggplot() + 
  # Construct the Heatmap Portion
  stat_density2d(data = filtered_df,
                 aes(x = position_long, y = position_lat, fill = after_stat(count)),
                 geom = 'tile',
                 contour = F,
                 n = 1000
                 ) +
  #Draw the map local trails
  geom_sf(data = roads, color = '#DDDDDD', alpha = .15) + 
  scale_fill_viridis_c(option = "B", guide = F) +
  ggthemes::theme_map() + 
  theme(
    panel.background = element_rect(fill = 'black'),
    plot.background = element_rect(fill = 'black')
  )

cowplot::ggdraw(p) + 
  labs(title = "My Trails Heatmap",
       caption = "An Adventure with scraping Strava data and mapping in R") +
  theme(panel.background = element_rect(fill = "black"),
        plot.background = element_rect(fill = 'black'),
        plot.title = element_text(color = "#DDDDDD",
                                  family = 'Calibri',
                                  #face = 'bold',
                                  size = 18),
        plot.caption = ggtext::element_markdown(color = '#DDDDDD',
                                    family = 'Calibri',
                                    size = 12))
```

### Make an interactive heatmap with leaflet
Also, just for fun, make a simple interactive leaflet map
```{r}
#| warning: false
run_map <- filtered_df  |> 
  leaflet()  |> 
  addTiles()  |>  # Add default OpenStreetMap map tiles
  addHeatmap(
    lng = ~position_long,
    lat = ~position_lat,
    blur = 20,    # Adjust blur and radius for desired look
    max = 0.5,
    radius = 15,
  )

# Display the map
run_map

```
