---
title: "Geoparquet in R"
date: "2024-04-17"
description: "Accessing an example geoparquet file in an S3 bucket and pulling data in only within a bounding box and for certain criteria"
categories: [geoparquet, R, arrow]
author: Marc Weber
---

Geoparquet extends the parquet file format and as nicely described in [this blog post by Kyle Barron](https://cloudnativegeo.org/blog/2024/12/interview-with-kyle-barron-on-geoarrow-and-geoparquet-and-the-future-of-geospatial-data-analysis/#:~:text=GeoParquet%20is%20a%20file%20format,single%20row%20of%20your%20data.) and provides a powerful new way to store and share geospatial data in a cloud-optimized format.  I've been using it for more and more of my spatial data, and below is just a quick example using Overture Maps buildings data and doing spatial and attribute filtering to subset the data prior to reading in.

### First open a connection to a cloud-hosted GeoParquet file
For this example we use Overture Maps buildings (public S3 bucket). We'll open a connection (but we are not actually reading it in yet)
```{r}
#| warning: false
#| message: false

library(arrow)
library(dplyr)
library(sf)
library(sfarrow)

# Connect to Overture S3 (anonymous, us-west-2)
bucket <- s3_bucket("overturemaps-us-west-2", anonymous = TRUE, region = "us-west-2")
ds_path <- bucket$path("release/2025-12-17.0/theme=buildings/type=building")
buildings_ds <- open_dataset(ds_path, format = "parquet")
# Inspect available columns to confirm tile partitioning
print(buildings_ds$schema$names)  # look for "z", "x", "y" as partition columns
bbox_field <- buildings_ds$schema$GetFieldByName("bbox")
```

### Look at a slice to see structure of data
```{r}
collect(
  buildings_ds %>%
    select(id, bbox) %>%
    slice_head(n = 5)
)
```

### Materialize a bbox as columns for inspection

```{r}
# Get the Field and its type
bbox_field <- buildings_ds$schema$GetFieldByName("bbox")
print(bbox_field)
print(bbox_field$type) 
```

### Prune the data
We define your query bbox in EPSG:4326 lon/lat
```{r}
xmin <- -123.32; ymin <- 44.52
xmax <- -123.20; ymax <- 44.63
bbox_sfc <- st_as_sfc(st_bbox(c(xmin = xmin, ymin = ymin, xmax = xmax, ymax = ymax), crs = 4326))


ds_pruned <- buildings_ds %>%
  filter(
    bbox$xmin <= xmax,      # feature xmin <= query xmax
    bbox$xmax >= xmin,      # feature xmax >= query xmin
    bbox$ymin <= ymax,      # feature ymin <= query ymax
    bbox$ymax >= ymin,      # feature ymax >= query ymin
    !is.na(height), height >= 4,
    !is.na(num_floors), num_floors >= 2
  ) %>%
  # Keep simple scalar columns plus geometry (avoid nested list/struct columns like 'names', 'sources', 'has_parts')
  select(id, geometry, height, num_floors, subtype, class)
```

### Materialize and convert to sf
```{r}
# Materialize and convert to sf
tbl <- collect(ds_pruned)
geom_sfc <- st_as_sfc(tbl$geometry, crs = 4326)
buildings_sf <- st_sf(tbl[, setdiff(names(tbl), "geometry")], geometry = geom_sfc)

# Optional: exact bbox cut in sf for precise geometry clipping
bbox_sfc <- st_as_sfc(st_bbox(c(xmin = xmin, ymin = ymin, xmax = xmax, ymax = ymax), crs = 4326))
buildings_bbox <- st_filter(buildings_sf, bbox_sfc)

mapview::mapview(buildings_bbox)
```





